---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# reasonabletools

<!-- badges: start -->
<!-- badges: end -->

This package provides the user with convenience functions for cleaning and summarizing data for use in Reasonable Potential Analyses (RPAs) and Water Quality-based Effluent Limitation (WQBEL) development. Examples of tasks one can do with reasonable tools include constructing composite parameter measurement (e.g., total PCBs) from congener water samples, finding maximum observed concentrations from partially censored datasets, and projecting maximum effluent concentrations (MECs). Methods for projecting MECs and computing effluent limitations are based on EPA's Technical Support Document for Water Quality-based Toxics Control (TSD). 

## Installation

You can install the current version of reasonabletools from [github](https://github.com/mattreusswig/reasonabletools) using devtools.

``` r
# If devetools is not installed, install it.
install.packages("devtools")

# Install reasonabletools from the github repository.
devtools::install_github("mattreusswig/reasonabletools")
```

## Usage

Functions in reasonabletools accept vectors as arguments--typically, a vector of numeric data representing concentration point measurements or detection limits, and a vector of character data indicating the detected/non-detect/detected-but-not-quantified status of each numeric value. 

```{r example}
library(reasonabletools)

## Create an example dataset
zinc <- data.frame(qual   = c(rep("", 10), rep("<", 10)),
                   result = 1:20,
                   stringsAsFactors = FALSE)
zinc

## Find observed MEC
find_mec(qual = zinc$qual, result = zinc$result)

```

### Non-Detects
By default, non-detect values and detected-but-not-quantified (DNQ or j-flag) values are both treated as non-detect and are indicated by either a "<", "nd", or "ND" character string. Any string in the non-detect qualifier vector which does not match these indicators is treated as a detected value. However, the non-detect indicators can be changed using the "nd" argument. In function output, "<" is always used to indicate a censored value and an empty string "" is used to indicate a detected value.

```{r cars}
## Create an example dataset with new custom non-detect indicator
zinc <- data.frame(qual   = c(rep("My Indicator", 20)),
                   result = 1:20,
                   stringsAsFactors = FALSE)
zinc

## Find observed MEC
find_mec(qual = zinc$qual, result = zinc$result, 
         nd = c("My Indicator"))

```

## Assumptions for CVs and Distributions

### Minimum Information Content for CV
Consistent with the TSD, reasonabletools requires a minimum level of information content in the dataset in order to estimate the Coefficient of Variation (CV) for a pollutant. There must be at least 10 observations and fewer than 80 percent of the dataset must be non-detect to calculate a CV value. If both those conditions are not met, a CV value of 0.6 is assumed and output.

### Treatment of Non-Detects
Consistent with the TSD, reasonabletools uses substitution methods when calculating mean and standard deviation values when non-detects are present. When non-detects are present, result values corresponding to non-detect qual values are multiplied by 0.5 prior to computing the mean and standard deviation.

### Distributional Assumptions
Consistent with the TSD, reasonabletools assumes a lognormal distribution is applicable to the data when performing effluent limitation computations.